Using VDMS to Search 100M Images: A Piece of Cake

### 2020-02-03 yfcc100m/paper/vldb2020/design.tex
More refs and other additions

\section{Design and Implementation}
\label{arch}

VDMS implements a client-server design.
The VDMS Server handles client requests concurrently and coordinates
request execution across the metadata and data components in order to return a
unified response. The metadata component is the Persistent Memory Graph
Database (PMGD). The data component is our custom library, the
Visual Compute Library (VCL). The VCL enables machine friendly enhancements to
visual data.
VDMS and its components are fully available as open source tools
\footnote{https://github.com/IntelLabs/\{vdms, pmgd, vcl\}}.

@@

\textbf{Persistent Memory Graph Database:}
Recent developments in persistent memory technologies
like 3D XPoint~\cite{IntelXPoint15}
promise storage elements  providing  nearly  the  speed  of  DRAM  and  the
durability of block-oriented storage. To provide an efficient storage
solution addressing the increasing popularity of connected data and
applications that benefit from graph like processing, we have designed
and implemented an in-persistent-memory graph database, PMGD, optimized
to run on a platform equipped with persistent memory.
PMGD provides a property graph model of data storage with the traditional
atomicity, consistency, isolation, and durability properties expected from
databases. The graph model makes it very suitable for the data model and
access patterns shown by visual metadata.
With its natural ability to extend the schema very
easily (due to the use of a property graph model),
we can support new developments in machine learning that can lead to
enhancements to existing metadata over time.
The specific details on how the data structures are optimized for persistent
memory, and the performance comparison with other graph databases is on-going
work that will be publish separately.

\textbf{Visual Compute Library:} The VCL was designed to provide an interface
through which users can interact with visual data. For traditional formats,
the interface is an abstraction layer over OpenCV. However, it also provides a
way to use novel formats that are better suited for visual analytics: a novel,
array-based lossless image format. This format is built on the array data
manager TileDB~\cite{TileDB} and is well suited for images that are used in
visual analytics.
The VCL currently provides limited support for videos but we are enhancing
its capabilities as part of our future work.
Feature vector support is provided through an implementation based
on high-dimensional sparse arrays, also using TileDB, which contains both
storage and search functionality over feature vectors.
In addition, the VCL provides a wrapper
for another high-dimensional index implementation,
Facebook's Faiss~\cite{faiss}.

\begin{comment}

\begin{figure*}[t!]
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.92\linewidth]{figures/query}
  \caption{Simple Metadata Query.}
  \label{fig:sfig1}
\end{subfigure}%
\begin{subfigure}{.52\textwidth}
  \centering
  \includegraphics[width=.94\linewidth]{figures/query2}
  \caption{Query involving visual transformations.}
  \label{fig:sfig2}
\end{subfigure}
\begin{subfigure}{\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{figures/results}
  \caption{Original image (left) and the result of query in (b)}
  \label{fig:sfig3}
\end{subfigure}
\caption{Snapshot of a Python notebook with sample queries}
\label{fig:query}
\end{figure*}

\end{comment}

\textbf{Request Server:}
Developers and users of machine learning frameworks and data science
applications favor simpler interfaces to access and process data and cannot
be expected to deal with two different ways of interacting with information (
metadata and visual data) instead of focusing on the algorithmic parts of their
pipelines.
VDMS takes care of coordinating client requests across the metadata and the
data as well as efficiently manages multiple clients through its Request
Server component, by implementing a JSON-based API.
It decomposes the command into
metadata and data requests, invokes the relevant calls behind the scene,
and returns a coherent response to the user after applying any additional
operations (explained in the next section).

\textbf{Client Library:}
A user application can use the VDMS API by defining metadata conforming to the
query protocol we have defined and passing along blobs. The client side VDMS
library provides a simple query function that accepts a JSON string with
commands and an array or vector of blobs.
Internally, the library wraps the query string and blob using Google Protobufs
and sends it to VDMS. It also receives a similarly wrapped response from VDMS
and returns it to the client. The responses require JSON parsing on the client
side for the metadata string that indicates how to interpret blobs.
VDMS provides Python and C++ client libraries with mainly
one function: a query that passes metadata as a JSON string and,
optionally, data as a vector of blobs. The VDMS client library manages the
rest.

\section{VDMS API}
\label{arch}
One of our goals with VDMS was to define an API that is easy to use.
Our API explicitly predefines certain
visual primitives associated with images, videos, and feature vectors. In
addition, while we use a graph database to store our metadata, the API is not
graph specific.
We have paid particular attention to hide the complexities of our internal
implementation and up-level the API to a JSON-based
API\footnote{https://github.com/IntelLabs/vdms/wiki/API-Description},
which is very popular across various application domains.
We understand that by defining a new JSON-based API we had to trade-off
expressiveness (compared to SPARQL or Gremlim) for the ability to design
native support for visual data, but we believe that, as development of VDMS
evolves, we will be able to achieve similar levels of expressiveness compared
to more mature query languages.
We have developed Python and C++ client
library to provide a simple query function that accepts a JSON string with
commands and an array or vector of blobs.

\begin{comment}
% All of the folliwng is commented out.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.9\columnwidth]{figures/features}
\caption{Feature Vectors natively supported in VDMS}
\label{fig:features}
\end{figure}

We use a medical imaging dataset as a driving example, and use it
here to showcase our API. More details about this use case are discussed in
Section~\ref{demo}.
Figure~\ref{fig:query} shows a snapshot from a Jupyter Notebook with two
simple queries.
Figure~\ref{fig:sfig1} shows a simple metadata query
where information about properties of patients matching some specific
characteristics (admitted with 85 years of age or more)
is used to constrain the search.
The returned JSON structure shows 2 patients found in the database.
Note that in this examples the queries are expressed as plain strings to
help the reader understand its structure. In real applications, the queries
are structured using the preferred method or library
for handling JSON structures (e.g. Python dictionaries or jsoncpp in C++).

Figure~\ref{fig:sfig2} shows a query involving visual transformations.
In this case, the query is looking for an image with a certain \textit{id}
value, and expects  VDMS to return the image twice: once with a thresholding
transformation (e.g., zero out all pixels less than a threshold), and the
second time after applying a threshold and then resizing it.
\ref{fig:sfig3} shows the image originally inserted in VDMS (left),
along with the returned images corresponding to the query in \ref{fig:sfig2}.
Adding new operations to VDMS is easy, as our software architecture
encapsulates all operations in VCL, and any new operation can use
or wrap around OpenCV, which is used internally.

\end{comment}
% End of Comment

Another key differentiating factor of VDMS is that it allows the creation of
indexes for high-dimensional feature vectors and the insertion of
these feature vectors associated with entities or visual objects.
Feature vectors are intermediate results of various machine
learning or computer vision algorithms when run on visual data.
These vectors can be labeled and classified to build search indexes,
and there are many in-memory libraries that are designed for
this task~\cite{flann, faiss}.
Using the VDMS API, users can manage feature vector indexes,
query previously inserted elements (images),
run a k-nearest neighbor search (knn) and express relationships
between existing images or feature vectors and
the newly inserted feature vectors.
By natively supporting feature vectors and knn,
VDMS allows out-of-the-box classification
functionalities for many applications.
Figure~\ref{fig:features} shows an example of how this functionality
can be integrated in a medical imaging analytics pipeline:
(1) A feature vector is extracted from an bounding
box in a brain image and labeled;
(2) the feature vector is inserted together
with all the associated metadata (type of cancer, for example)
and a link to the original image.
(3) A new feature vector is extracted from a new image,
(4) a query to VDMS is issued to classify that feature vector based on the
indexed features in VDMS, and
(5) VDMS can respond to the query with the label
associated to that feature vector.
Note that, even if the example is based on the medical images used for the
performance evaluation, this methodology is applicable in many other contexts
and use cases, such as face detection and matching.
More details about the JSON API for this functionality can be found on
our Github wiki page.

>>>>

\section{VDMS Desing and Implementation}
\label{arch}

In this section, we briefly describe VDMS design and implementation,
but most of it is already cover in previous work \cite{vdms-nips}.
Figure \ref{fig:arch} depicts the high-level architecture of VDMS.
VDMS implements a client-server architecture that handles client
requests concurrently and coordinates query execution across
the metadata and data components in order to return a unified response.

The metadata component is the Persistent Memory Graph
Database (PMGD). The (visual) data component is our Visual Compute Module.
The Visual Compute Module enables machine friendly enhancements to
visual data, exposing high-level abstractions to the Request Server for
dealing with a variety of images and video formats (through OpenCV),
and different methods for indexing for Feature Vectors
(including Facebook's Faiss \cite{faiss}, TileDB \cite{TileDB}).
VDMS and its components are fully available open source
\footnote{https://github.com/IntelLabs/\{vdms, pmgd\}}.
We briefly describe each of the main components as follows:

@@

\textbf{Persistent Memory Graph Database:}
We use PMGD to provide an efficient storage
solution addressing the increasing popularity of connected data and
applications that benefit from graph like processing, we have designed
and implemented an in-persistent-memory graph database, PMGD, optimized
to run on a platform equipped with persistent memory.
PMGD provides a property graph model of data storage with the traditional
atomicity, consistency, isolation, and durability properties expected from
databases. The graph model makes it very suitable for the data model and
access patterns shown by visual metadata.
With its natural ability to extend the schema very
easily (due to the use of a property graph model),
we can support new developments in machine learning that can lead to
enhancements to existing metadata over time.
The specific details on how the data structures are optimized for persistent
memory, and the performance comparison with other graph databases is on-going
work that will be publish separately.
PMGD has been design and optimized for persistent memory technologies
like 3D XPoint~\cite{IntelXPoint15}, which
promise storage providing nearly the speed of DRAM and the
durability of block-oriented storage.

\textbf{Visual Compute Module:} This module was designed to provide
an abstraction layer for interacting with with visual data.
For traditional formats (jpg, png, tiff, mp4, etc),
the interface is an abstraction layer over OpenCV. However, it also provides a
way to use novel formats that are better suited for visual analytics: a novel,
array-based lossless image format. This format is built on the array data
manager TileDB~\cite{TileDB} and is well suited for images that are used in
visual analytics.
This module also provides support for videos, enabling operations like
encoding/decoding/transcoding and resize as part of VDMS interface.
Feature vector support is provided through an implementation based
on high-dimensional sparse arrays, also using TileDB, which contains both
storage and search functionality over feature vectors.
In addition, the VCL provides a wrapper
for another high-dimensional index implementation,
Facebook's Faiss~\cite{faiss}.

\textbf{Request Server:}
Developers and users of machine learning frameworks and data science
applications favor simpler interfaces to access and process data and cannot
be expected to deal with two different ways of interacting with information
(metadata and visual data) instead of focusing on the
algorithmic parts of their pipelines.
VDMS takes care of coordinating client requests across the metadata and the
data as well as efficiently manages multiple clients through its Request
Server component, by implementing a JSON-based API.
It decomposes the command into
metadata and data requests, invokes the relevant calls behind the scene,
and returns a coherent response to the user after applying any additional
operations (explained in the next section).

\textbf{Client Library:}
A user application can use the VDMS API by defining metadata conforming to the
query protocol we have defined and passing along blobs. The client side VDMS
library provides a simple query function that accepts a JSON string with
commands and an array or vector of blobs.
Internally, the library wraps the query string and blob using
Google Protobufs \cite{protobufs} and sends it to VDMS.
It also receives a similarly wrapped response from VDMS
and returns it to the client. The responses require JSON parsing on the client
side for the metadata string that indicates how to interpret blobs.
VDMS provides Python and C++ client libraries with mainly
one function: a query that passes metadata as a JSON string and,
optionally, data as a vector of blobs. The VDMS client library manages the
rest.

\textbf{VDMS API:}
VDMS API is easy to use and explicitly predefines certain
visual primitives associated with images, videos, and feature vectors. In
addition, while we use a graph database to store our metadata, the API is not
graph-specific.
Authors have paid particular attention to hide the complexities of our internal
implementation and up-level the API to a JSON-based
API\footnote{https://github.com/IntelLabs/vdms/wiki/API-Description},
which is very popular across various application domains.
By defining a new JSON-based API, there is a trade-off between
expressiveness (compared to SPARQL or Gremlim, or even SQL) and
the ability to natively support for visual data operations.
However, it should be possible to achieve similar levels of
expressiveness compared to more mature query languages over time.
The current front-end available are a Python and C++ client
library to provide a simple query function that accepts a JSON string with
commands and an array or vector of blobs.

While there are a number of big-data frameworks~\cite{spark, hadoop}, systems
that can be used to store metadata~\cite{memsql, vertica}, and systems that
manipulate a specific category of visual data~\cite{scidb, rasdaman}, VDMS can
be distinguished from them on the following aspects:

\begin{itemize}
\item {\em Design for analytics and machine learning}: by targeting
visual data for use cases that require manipulation
of visual information and associated metadata,
\item {\em Ease-of-use}: By defining a common API that allows applications to
combine their complex metadata searches with operations on resulting visual
data, and together with full support for feature vectors, VDMS goes beyond the
traditional SQL or OpenCV level interfaces that do one or the other. Given our
focus on enabling machine learning, we also provide a client API in Python.
\item {\em Performance}: We show how a unified system such as VDMS can
outperform an ad-hoc system constructed with well-known discrete components.
Because of the capabilities we have built into VDMS, it handles complex
queries significantly better than the ad-hoc system without compromising the
performance of simple queries.
\end{itemize}

### 2020-02-03 yfcc100m/paper/vldb2020/eval.tex
More refs and other additions

For all our experiments, we use 2 servers, one hosting VDMS server and
another hosting MySQL server. Both servers have a dual-socket
Intel\textsuperscript{\textregistered}
Xeon\textsuperscript{\textregistered} Platinum 8180 CPU @ 2.50GHz (Skylake),
each CPU with 28 physical cores with multithreading enabeled,
for a total of 112 logical cores per server.
The server hosting MySQL has 256GB of DDR4 DRAM, while the server hosting VDMS
has 64GB of DDR4 DRAM. Both servers run Ubuntu 16.04.
We use MySQL version XXXXX.

As a baseline, we implemented a similar database system comprised of a
combination of widely available and out-of-the-box components: MySQL Server 5.7,
Apache Web Server, and OpenCV.
Figure \ref{figure:systems} shows a logical view of the difference between the
interaction of the client application (which wants to retrieve metadata and
imaged) with VDMS (left) and the baseline (right).
It is worth noting that the images are stored in a shared repository
(ext4 filesystem on a RAID 6 configuration of 16TB) that both Apache WebServer
and VDMS have direct access. In the case of MySQL, metadata is stored in an
attached SSD disk. Even if VDMS has native support for Optane Persistent Memory,
we do not use it in this experiment because of fairness of comparison w.r.t
MySQL. The benefits of Persistent Memory on metadata operations is left
for another paper, and outside the scope of this evaluation.
For this experiment then, we simply use a similar attached SSD disk to store
metadata (even if VDMS Graph store is designed for PM, it can deliver good
performance when using SSDs directly).

@@

>>>>

For all our experiments, we use 2 servers, one hosting VDMS server and
another hosting MySQL server. Both servers have a dual-socket
Intel\textsuperscript{\textregistered}
Xeon\textsuperscript{\textregistered} Platinum 8180 CPU @ 2.50GHz (Skylake),
each CPU with 28 physical cores with multithreading enabeled,
for a total of 112 logical cores per server.
The server hosting MySQL has 256GB of DDR4 DRAM, while the server hosting VDMS
has 64GB of DDR4 DRAM. Both servers run Ubuntu 16.04.

As a baseline, we implemented a similar database system comprised of a
combination of widely available and out-of-the-box components: MySQL Server 5.7,
Apache Web Server 2.4.18, and OpenCV 3.3.
Figure \ref{figure:systems} shows a logical view of the difference between the
interaction of the client application (which wants to retrieve metadata and
imaged) with VDMS (left) and the baseline (right).
It is worth noting that the images are stored in a shared repository
(ext4 filesystem on a RAID 6 configuration of 16TB) that both Apache WebServer
and VDMS have direct access. In the case of MySQL, metadata is stored in an
attached SSD disk. Even if VDMS has native support for Optane Persistent Memory,
we do not use it in this experiment because of fairness of comparison w.r.t
MySQL. The benefits of Persistent Memory on metadata operations is left
for another paper, and outside the scope of this evaluation.
For this experiment then, we simply use a similar attached SSD disk to store
metadata (even if VDMS Graph store is designed for PM, it can deliver good
performance when using SSDs directly).

@@

Another key differentiating factor of VDMS is that it allows the creation of
indexes for high-dimensional feature vectors and the insertion of
these feature vectors associated with entities or visual objects.
Feature vectors are intermediate results of various machine
learning or computer vision algorithms when run on visual data.
These vectors can be labeled and classified to build search indexes,
and there are many in-memory libraries that are designed for
this task~\cite{flann, faiss}.
Using the VDMS API, users can manage feature vector indexes,
query previously inserted elements (images),
run a k-nearest neighbor search (knn) and express relationships
between existing images or feature vectors and
the newly inserted feature vectors.
By natively supporting feature vectors and knn,
VDMS allows out-of-the-box classification
functionalities for many applications.

% Figure~\ref{fig:features} shows an example of how this functionality
% can be integrated in a medical imaging analytics pipeline:
% (1) A feature vector is extracted from an bounding
% box in a brain image and labeled;
% (2) the feature vector is inserted together
% with all the associated metadata (type of cancer, for example)
% and a link to the original image.
% (3) A new feature vector is extracted from a new image,
% (4) a query to VDMS is issued to classify that feature vector based on the
% indexed features in VDMS, and
% (5) VDMS can respond to the query with the label
% associated to that feature vector.
% Note that, even if the example is based on the medical images used for the
% performance evaluation, this methodology is applicable in many other contexts
% and use cases, such as face detection and matching.
% More details about the JSON API for this functionality can be found on
% our Github wiki page.

### 2020-02-03 yfcc100m/paper/vldb2020/vdms-vldb2020.tex
More refs and other additions

\documentclass{vldb}
\usepackage{graphicx}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
\usepackage{hyperref}       % hyperlinks
\usepackage{graphicx}
\usepackage{comment}
\usepackage{multicol}
\usepackage{framed}
\usepackage{subcaption}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}

@@

\begin{abstract}
Data scientists spend most of their time dealing with data preparation,
rather than doing what they know best:
build machine learning models and algorithms to solve problems.
The problem grows as the scale of the data that is used for analysis or training
grows. In this paper, we demonstrate how VDMS (Visual Data Management System)
can be used to simplify the data preparation process and we show how we gain
in efficient just because we are using a tool that was design for the job.
To demonstrate this, we use the largest available public dataset (YFCC100M),
with 99+ million images, 800+ thousand videos plus add-ons that include
machine-generated tags and 4K dimensional feature vectors,
a total of ± 13TB of data.
VDMS differs from existing large scale photo
serving, video streaming, and textual big-data management systems due to
its primary focus on supporting machine learning and
data analytics pipelines that use visual data
(images, videos, and feature vectors), treating these as first class citizens.
We demonstrate how VDMS outperforms well-know and widely used tools for
data access by a 10x factor in certain queries, while simplifying the process
of data preparation and data access, and providing functionalities non-existent
in alternative options.
\end{abstract}

@@

We have to acknowledge Philip, Vishakha, Nilesh, and any other reviewer.

% \balance

@@

\begin{appendix}
You can use an appendix for optional proofs or details of your evaluation which are not absolutely necessary to the core understanding of your paper.

\section{Final Thoughts on Good Layout}
Please use readable font sizes in the figures and graphs. Avoid tempering with the correct border values, and the spacing (and format) of both text and captions of the PVLDB format (e.g. captions are bold).

At the end, please check for an overall pleasant layout, e.g. by ensuring a readable and logical positioning of any floating figures and tables. Please also check for any line overflows, which are only allowed in extraordinary circumstances (such as wide formulas or URLs where a line wrap would be counterintuitive).

Use the \texttt{balance} package together with a \texttt{\char'134 balance} command at the end of your document to ensure that the last page has balanced (i.e. same length) columns.

>>>>

\documentclass{vldb}
\usepackage{graphicx}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
\usepackage{hyperref}       % hyperlinks
\usepackage{graphicx}
\usepackage{comment}
\usepackage{multicol}
\usepackage{framed}
\usepackage{subcaption}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{mathtools}

@@

\begin{abstract}
Data scientists spend most of their time dealing with data preparation,
rather than doing what they know best:
build machine learning models and algorithms to solve problems.
The problem grows as the scale of the data that is used for analysis or training
grows. In this paper, we demonstrate how VDMS (Visual Data Management System)
can be used to simplify the data preparation process and we show how we gain
in efficiency just because we are using a tool that was design for the job.
To demonstrate this, we use one of the largest available
public dataset (YFCC100M),
with 99+ million images, 800+ thousand videos plus add-ons that include
machine-generated tags and 4K dimensional feature vectors,
a total of $\sim$13TB of data.
VDMS differs from existing large scale photo
serving, video streaming, and textual big-data management systems due to
its primary focus on supporting machine learning and
data analytics pipelines that use visual data
(images, videos, and feature vectors), treating these as first class citizens.
We demonstrate how VDMS outperforms well-know and widely used tools for
data access by a $\sim$10x factor in certain queries,
while simplifying the process
of data preparation and data access, and providing functionalities non-existent
in alternative options.
\end{abstract}

@@

We have to acknowledge Philip and Vishakha for help in fixing PMGD bug,
Nilesh for his help/input, and any other reviewer.

@@

\balance
\begin{appendix}
You can use an appendix for optional proofs or details of your evaluation which are not absolutely necessary to the core understanding of your paper.

% \section{Final Thoughts on Good Layout}
% Please use readable font sizes in the figures and graphs. Avoid tempering with the correct border values, and the spacing (and format) of both text and captions of the PVLDB format (e.g. captions are bold).

% At the end, please check for an overall pleasant layout, e.g. by ensuring a readable and logical positioning of any floating figures and tables. Please also check for any line overflows, which are only allowed in extraordinary circumstances (such as wide formulas or URLs where a line wrap would be counterintuitive).

% Use the \texttt{balance} package together with a \texttt{\char'134 balance} command at the end of your document to ensure that the last page has balanced (i.e. same length) columns.
